The role of cross-linguistic lexical similarity on bilingual word acquisition

### 2020-06-30 Manuscript/trajectories_manuscript.Rmd
First version of the Introduction and Methods, and second version of the ICIS 2020 poster

affiliation:
  - id            : "1"
    institution   : "Center for Brain and Cognition, Universitat Pompeu Fabra"

authornote: |
  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.

  Enter author note here.

abstract: |
  Previous literature on early vocabulary has been mostly committed to exploring how the receptive and productive vocabulary size changes with age, and its relationship with toddlers’ performance on language tasks (e.g. Fernald, Swingley, & Pinto, 2001; Fernald, Perfors, & Marchman, 2006). The content of the developing lexicon has remained relatively unexplored. More recently, studies have focused on the characterisation of the developmental trajectory of individual words, reporting an earlier age of acquisition for words with high frequency, concreteness, and phonological neighbourhood density (e.g. Braginsky, Yurovsky, Marchman, & Frank, 2019; Jones & Brandt, 2019). Such item-level analysis allows not only to predict the age of acquisition of specific words, but also to shed light on the cognitive processes that underlie early word learning. This approach is particularly interesting for investigating the potential impact of acquiring two languages simultaneously. A recent study by Floccia et al. (2018) compared vocabulary sizes of 24-month-old children learning British English, together with an additional language from a pool of 13 diverse languages. They found larger productive vocabulary sizes of toddlers learning two languages that are phonologically similar (see for a similar approach Bosch & Ramon-Casas, 2014). The mechanisms underlying this effect remain unknown. One possibility is that the similarity between the two languages speeds the acquisition of form-similar translation equivalents. The aim of this study is to perform an item-wise analysis on infant’s vocabulary contents. We developed an online tool to collect parental reports of receptive and productive vocabularies from children learning Catalan and/or Spanish. We expect that phonological overlap between translation equivalents will predict earlier age of acquisition. If this effect is driven by the phonological overlap between translation equivalents, cognate pairs should be acquired closer in time than non-cognates. For instance, the translation of cat /gato/ in Spanish and /gat/ in Catalan should be learnt at approximately the same age. This should not necessarily happen for the translations of dog, /pero/ and /gos/. We analyse the time elapsed between the acquisition of a word in one language, and its translation in the other.
  We present preliminary data (data collection is ongoing) on receptive vocabulary of 90 monolingual and bilingual toddlers aged 18 to 36 months. We obtained parental responses to 230 pairs of Catalan-Spanish translation equivalents, resulting in a total of 41490 responses. Words that yielded a phonologically similar translation equivalent were more likely to be present in toddlers’ receptive vocabulary at all ages, and were acquired earlier, even when accounting for the effect of word frequency. Further analysis will explore the distance in time between the acquisition of pairs of translation equivalents using a more detailed measure of phonological overlap across translation equivalents and taking into account cross-individual and cross-item variability.

@@

keywords          : "word acquisition, vocabulary, bilingualism, lexical similarity"
wordcount         : "X"

bibliography      : ["../References/BiLexicon.bib"]

@@

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

@@

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  cache.extra = knitr::rand_seed,
  results = "asis",
  out.width = "60%",
  fig.align = "center"
)
```

```{r prepare, include=FALSE}
# load packages
library(papaja)
library(knitr)

library(magrittr)
library(tibble)
library(dplyr)
library(forcats)
library(stringr)
library(readxl)
library(tidyr)
library(lubridate)

library(ggplot2)
library(ggridges)
library(patchwork)
library(wesanderson)

library(truncnorm)
library(data.table)
library(here)

# load/create functions
source(here("R", "functions.R"))

# set params
age_bins <- c("12-14", "14-16", "16-18", "18-20", "20-22", "22-24", "24-26", "26-28", "30-32")

```{r prepare_data, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

# prepared data
dat_items <- fread(here("Data", "04_prepared.csv"), na.strings = c("")) %>%
  as_tibble() %>%
  drop_na(language)

# load data
dat <- fread(here("Data", "02_merged.csv"), na.strings = "") %>%
  filter(source %in% c("formr1", "formr2", "formr_short"),
         lp %in% c("Monolingual", "Bilingual"),
         age_bin %in% age_bins) %>%
  as_tibble() %>%
  mutate_at(vars(date_birth, time_stamp), as_date)

# import data
wordbank <- fread(here("Data", "05_wordbank.csv"))
wordbank_priors <- fread(here("Data", "05_wordbank-priors.csv"))
mid_prior <- round(wordbank_priors$estimate[wordbank_priors$term=="mid"], 2)
mid_steep <- round(wordbank_priors$estimate[wordbank_priors$term=="steep"], 2)

# load results
bayes_factor <- readRDS(here("Results", "bayes_factor.rds"))
loo_comp <- readRDS(here("Results", "loo.rds"))
fit0_production <- readRDS(here("Results", "fit0_productive.rds"))
fit1_production <- readRDS(here("Results", "fit1_productive.rds"))
fit1_priors_production <- readRDS(here("Results", "fit1_prior_productive.rds"))

@@

One of the main challenges bilingual infants and toddlers face is learning two distinct set of words (one for each language) that partially overlap in sound and meaning. Previous studies have reported that bilingual children know fewer words than their monolingual counterparts when only language is considered (e.g., English monolinguals know more words in English than English-Spanish bilinguals). When taking both languages into account, bilingual children between seem to know, at least, as many words as monolinguals do [@benzeev1977; @doyle1978; @rosenblum1983; @fernandez1992; @pearson1994; @bialystok2010; @blom2019]. Tough, not all bilinguals show asimilar developmental trajectory of lexical acquisition.

Recent studies have capitalised on the role of the similarity between the specific pair of language the infant is learning. @floccia2018 analysed vocabulary scores across of 24-month-old toddlers learning English and an additional language. Toddlers learning languages sharing a large amounth of cognates (e.g., English-German) showed larger vocabulary sizes than those learning less languages sharing less cognates (e.g., English-Chinese). @blom2019 extended these results to children aged three to 10 years. These results point to the possibility that the overlap between the word inventories infants are acquiring impacts their trajectory of lexical aquisition. However, none of these studies provide an account for the mechanisms involved in this facilitatory effect of language driven by similarity across languages. The aim of this study is to explore two scenarios in which this effect may be taking place:

1) In a first scenario, it is the phonological overlap between language pairs, and not the amount of cognates they share, what boosts the vocabulary growthof bilingual toddlers. Languages sharing a lot of cognates tend to also share a lot of phonemic categories. It is possible that having to learn fewer phonemic categories makes it easier for toddlers to acquire words. If this is true, toddlers learning two languages with similar phonemic inventories should show larger vocabulary sizes than those learning languages that barely ovelap phonologically. Moreover, this facilitation effect should be reflected in the developmental trajectory of all the items included in a vocabulary inventory, independently of their cognate status.

2) In a second scenario, Words that are form similar It is the amount of cognates language pairs share that boosts vocabulary growth in bilingual toddlers. If this is true, toddlers may acquire cognates earlier

To our knowledge, there are no previous studies that have investigated the effect of cognateness on the acquisition trajectories of individual word forms. Braginsky et al. (2019) used data from MacArthur-Bates CDIs from multiple languages to explore what properties are associated with an earlier acquisition. The found that frequency, concreteness, and [...] significantly predicted an earlier age of acquisition. We plan to extend these analyses to the case of bilinguals.

We have developed a fully-online vocabulary inventory in Catalan and Spanish that is more exhaustive than the MacArthur-Bates CDI, including a larger number of items. All participants filled the questionnaire it both languages, thus providing responses to pairs of translation equivalents.

In this study, we examine the role of cognateness on age of acquisition by modelling the probability of parents reporting that the participant is able to understand, say and understand on each item in the questionnaire. We will test the effect the participants' language profile (monolingual vs. bilingual), and its cognate status on the developmetnal curves of each item.

We will analyse the developmental trajectories of individual words in monolinguals and bilinguals learning Catalan and Spanish (two phonologically close languages sharing a great deal of cognates), between 14 and 30 months of age. We will test whether cognate words are learnt earlier than non-cognate words, taking into account wether they are part of the most- o less-dominant language of the infant, as well as their frequency, concreteness, and number of syllables. We have deloped an on-line questionnaire that entails a short language exposure questionnaire, asks for some demographic information, and presents parent with a list of ~800 words in each language, for which parent report whether their child understands, says, or doesn't understand and say, in each item.

We will fit a Bayesian multilevel model on the item responses, modelling the probability of parents reporting that a given word is acquired by the infant, including language profile (monolingual vs. bilingual), item dominance (dominant language vs. non-dominant language), frequency, concreteness, and number of syllables. We will include *meaning* as grouping variable. We will fit a sigmoidal curve [see @mayor2008] on each item, defined by the steepness of the developmental curve, and the mid-point (the poinnt at which the developmental curve is steepest, which will be considered the point of acquisition). Priors will be derived from previous literature on the subject.

NOTES:

Evidence that bilingual lexica barely overlap [@leopold1948; @taeschner1978].

Studies with infants from 8 to 30 months of age show weaker evidence for this claim [@pearson1993; @dehoouwer2014], pointing to the posibility that differences between bilingual and monolingual vocabulary sizes may be dependent on maturational factors.

sugested that the developmental trajectory of lexical aquisition varies across two dimensions: the language profile of the infant, and the joint proporties of the specific language pair the infant is learning. Regarding the first, @pearson1994 found thant infants exposed to a more balanced

Of special interest is the fact than cognates (form-similar translation equivalents) tend to be overrepresented in the early bilingual vocabulary [@bosch2014].

The similarity between both languages may play a role in bilingual lexical aquisition.

## Hypotheses

Hypothesis 1
:  The boost effect of language similarity is driven by cognates being acquired earlier than non-cognates. We predict that age of acquisition of cognates will be ealier thant that of non-cognates. This effect should only be present in bilinguals.

Hypothesis 2
:  That this effect is driven by the word-forms in one language scaffolding the acquisition of their translation equivalents via parallel activation. We predict that the difference in time between the acquisition pairs of translation equivalents is shorter in cognates than in non-cognates. This effect will only be present in bilinguals.

@@

Data from `r dat$id_db %>% unique %>% length() %>% printnum()` participants from the different cities in Spain were invited to participate through social media. Families in the Metropolitan area of Barcelona were also recruited at birth. All families participated voluntarily. Data were collected between `r min(dat$time_stamp, na.rm = TRUE) %>% format(., "%dth %B, %Y")` and `r max(dat$time_stamp, na.rm = TRUE) %>% format(., "%dth %B, %Y")`, using the BiLexicon inventory. This study was approved by the Comitè d'Ètica de la Investigació amb Medicaments (CEIm) from Hospital del Mar (Barcelona, Spain), code XXXXXXXXX.

```{r participants, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

dat %>%
  group_by(dominance, sex) %>%
  summarise(n = n(),
            age_mean = mean(age, na.rm = TRUE),
            age_sd = sd(age, na.rm = TRUE),
            doe_catalan_mean = mean(doe_catalan, na.rm = TRUE),
            doe_catalan_sd = sd(doe_catalan, na.rm = TRUE),
            doe_spanish_mean = mean(doe_spanish, na.rm = TRUE),
            doe_spanish_sd = sd(doe_spanish, na.rm = TRUE)) %>%
  ungroup() %>%
  kable(digits = 2,
        col.names = c("Dominance", "Sex", "N", "M_Age", "SD_Age", "M DOE Cat", "SD_DOE_Cat",  "M_DOE_Spa", "SD_DOE_Spa"),
        format.args = list(big.mark = ",", na.string = "-"),
        align = "c",
        caption = "Sample demographics and overall language profile. Dominance = Language to which the participant is exposed the most, N = sample size, M age = Mean age, SD age = standard deviation of age, M DOE Cat = mean degree of exposure to Catalan, SD DOE Cat = standard deviation of exposure to Catalan, M DOE Spa = mean degree of exposure to Spanish, SD DOE Spa = standard deviation of exposure to Spanish.")

@@

* *Location*
* *Age*
* *Sex*
* *Language profile*: : Linguistic profiles were assessed using the Language Exposure Questionnaire (LEQ) [@bosch2001]. This questionnaire estimates de degree of exposure (DOE) to each language via a 10 minutes long interview with the parents. LEQs were administered inmediately before or after the testing session at the lab. We here considered only exposure to Catalan and Spanish. Participants with >10% exposure to a third language other than Catalan or Spanish were excluded.
* *SES*

### Exclusion criteria

* Language disorders

### Questionnaires

#### BiLexicon 1.0

#### BiLexicon 2.0

#### BiLexicon Short

This questionnaire was developed as a short version of the BiLexicon 1.0 and BiLexicon 2.0 inventories. Given the large amount of items to be answered to, we restricted the use of such inventories to participants that were to participate presentially in one of the studies run in the lab at that time. These participants only filled one of the questionnaires, and did so only once.

The aim of developing a short version of this questionnaire (BiLexicon Short) was to provde a more on-line friendly format to families that would participate remotely exclusively, or that would fill the questionnaire more than once (e.g., in longitudinal studies). We divided the items of each language in the pool into four versions named A, B, C, and D. Items in each category (e.g., animals, auxiliary words) were randomly assigned one of the versions. Thos categories that did not entail more than 16 items (thus resulting in less than four items after dividing it in ofur lists), were not divided (all of their items were included in all versions).

When filling the questionnaire for the first time, participants were randomly assigned one version. Each familiy filled only one of the versions, thus reducing the number of items to be answered. To ensure that longitudinal participants would provide longitudinal data in the questionnaire, we made sure that participants were always assigned the same version of this questionnaire. Words that were used as part of the trial lists in the experiment families would participate in were present across all versions.

### Item inclusion criteria

We gathered data from `r dat_items %>% distinct(language, meaning) %>% group_by(language) %>% summarise(n = n()) %>% pull(n) %>% min() %>% printnum(round = 0)` distinct translation equivalents across Spanish (`r dat_items %>% distinct(language, meaning) %>% group_by(language) %>% summarise(n = n()) %$%  n[2] %>% printnum()` word forms) and Catalan (`r dat_items %>% distinct(language, meaning) %>% group_by(language) %>% summarise(n = n()) %$% n[1] %>% printnum()` word forms). The unequal number of word-orms across both languages is due to that some tanslation equivalents did not entail a one-to-one-corespondence: some items had more than one translation equivalent in the other language (e.g. the Catalan word form *walking*, *anar*, can be translated as both *caminar* [to walk] or *ir* [to go]).

We included items from the following categories: Action words, adventures, animals, body parts, clothes, color, descriptive words, food and drink, furniture and rooms, household items, on-line, outside, parts of animals, parts of things, people, pronouns, quantifiers, question words, time, toys, and vehicles. We discarded the following categories: adverbs, auxiliary words, connectives, interjections and games and routines.

We excluded items that enailed more than one word (e.g., *barrita de cereales* [cereal bar]).

### Item properties

#### Frequencies

We retrieved frequencies from SUBTLEX-ESP [@cuetos2009] and SUBTLEX-CAT [@ferre2013] for Spanish and Catalan words, respectively. Word frequency was calculated as the Zipf score of the word [@zipf1949; @vanheuven2014].

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

include_graphics(here("Figures", "pool_freq_cognateness.png"))

>>>>

affiliation:
  - id            : "1"
    institution   : "Center for Brain and Cognition, Universitat Pompeu Fabra, Barcelona, Spain"

authornote: |
  "Campus de Ciutadella, Universitat Pompeu Fabra, 08005, Barcelona, Spain

abstract: |
 Bilinguals face the challenging task of learning words from languages with overlapping phonologies. Floccia et al. (2018) reported larger vocabulary sizes for 24-month-old bilinguals that were learning languages that shared a greater amount of cognates (e.g., English-Dutch). The mechanisms underlying this effect remain unknown. We explore two compatible scenarios. First, we test whether cognates are learnt earlier than non-cognates. This would account for the difference in vocabulary size associated to the amount of shared cognates across languages. Second, we explore the possibility that the word-forms of one language interact with those form the other language, scaffolding the acquisition of their translation equivalents when their phonologies overlap. This mechanism, in line with the parallel activation account of bilingual speech perception, would provide a plausible explanation to why cognates are acquired ealier by bilinguals. We developed an online tool to collect parental reports of receptive and productive vocabularies from children learning Catalan and/or Spanish, and present data on receptive and productive vocabulary of bilingual toddlers aged 12 to 34 months.

@@

keywords          : "lexical acquisition, vocabulary, bilingualism"
wordcount         : "X"

bibliography      : ["../References/trajectories.bib"]

@@

documentclass     : "apa6"
classoption       : "man"
output            :
  word_document:
    reference_docx: trajectories_template.docx

---

@@

knitr::opts_chunk$set(
	echo = FALSE,
	fig.align = "center",
	message = FALSE,
	warning = FALSE,
	cache.extra = knitr::rand_seed,
	out.width = "80%",
	results = "asis"
)
```

```{r prepare, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

# load packages
library(papaja)
library(knitr)

library(magrittr)
library(tidyverse)
library(dplyr)
library(forcats)
library(stringr)
library(readxl)
library(lubridate)
library(janitor)

library(ggridges)
library(patchwork)
library(gt)

library(data.table)
library(here)

# load/create functions
source(here("Code", "R", "functions.R"))

# set params
bins <- c("14-16", "16-18", "18-20", "20-22", "22-24", "24-26", "26-28", "28-30", "30-32", "32-34", "34-36")

# import data 
dat_studies <- fread(here("Data", "00_studies.csv")) %>% distinct(q_version, language, q_items)
dat_items <- read_xlsx(here("Data", "01_pool.xlsx")) %>%
  group_by() %>%
  drop_na(language) %>%
  mutate_at(vars(A, B, C, D, cognate_rater1, include, frequency), as.numeric) %>%
  mutate_at(vars(A, B, C, D, cognate_rater1, include), as.logical)
dat_participants <- list.files(here("Data", "Logs"), pattern = "logs", full.names = TRUE) %>%
  last() %>%
  fread(na.strings = "") %>%
  filter(age_bin %in% bins,
         completed,
         lp == "Bilingual")
dat_raw <- fread(here("Data", "02_merged.csv"), na.strings = "") %>%
  filter(lp %in% c("Bilingual"),
         age_bin %in% bins) %>%
  as_tibble() %>%
  mutate_at(vars(date_birth, time_stamp), as_date)

dat_wordbank <- fread(here("Data", "05_wordbank.csv"))
dat_wordbank_priors <- fread(here("Data", "05_priors-wordbank.csv"))

@@

Bilingual toddlers must learn two distinct set of words (one for each language) that partially overlap in sound and meaning. Previous studies have reported that bilingual children know fewer words than their monolingual counterparts when only language is considered (e.g., English monolinguals know more words in English than English-Spanish bilinguals). When taking both languages into account, bilingual children  seem to know, at least, as many words as monolinguals do [@ben-zeev1977; @doyle1977some; @rosenblum1983word; @fernandez1992; @pearson1994; @bialystok2010; @hoff2012; @core2013; @hoff2014; @blom2019; but see @pearson1993; @houwer2014]. Not all bilinguals show a similar developmental trajectory of lexical acquisition. Recent studies have capitalised on the role of the similarity between the specific pair of languages the infant is learning. One way of measuring language similarity is by estimating the amount of cognates (i.e., form-similar translation equivalents) they share. @floccia2018 explored the impact of this form of similarity on vocabulary sizes from a large sample of 24-month-old toddlers learning English and an additional language.  Toddlers learning languages that shared a large amount of cognates (e.g., English-German) showed larger vocabulary sizes than those learning languages that shared fewer cognates (e.g., English-Spanish). @blom2019 extended these findings to children aged three to 10 years. These results suggest that the phonological overlap between the word inventories infants are acquiring impacts their trajectory of lexical aquisition.

How does cognateness influence word learning? @floccia2018 point to the possibility that parallel activation during speech perception may play a role in this cognate facilitation effect. The parallel activation principle suggests that bilinguals activate both languages in parallel during speech perception and production, even when only one of them is being used. This hypothesis can only be understood under the more general account of language non-selectivity of lexical processing, which states that translation equivalents are co-activated at the lexical level [e.g., @de1991lexical; @costa2000; @dijkstra2005; @thierry2007; @hoshino2008; @yudes2010; @bobb2020; see @kroll2013 for review]. A growing body of experimental evidence has been given shape to different formal models of bilingual lexical processing, like the Revised Hierarchical model [@kroll1994; @kroll2010], the Bilingual Interactive Activation model [BIA; @dijkstra1998bia], and its successor BIA+ [@dijkstra2002architecture], the Bilingual Language Interaction Network for Comprehension of Speech model [BLINCS; @shook2013] and, more recently, Multilink [@dijkstra2019], which represents an effort for integrating the former to account for language-non selective activation during both comprehension and production. However, these models do not address word learning during language acquistion.

It is feasible that parallel activation plays a role in vocabulary acquisition during the first years of life. On top of Floccia et al.'s findings, other studies have provided experimental evidence of parallel activation during speech perception in bilingual toddlers [@vonholzen2012; @vonholzen2019]. Yet, it is unclear how parallel activation influences word learning at these ages. What mechanisms are engaged, and how do they relate to vocabulary size? We contemplate two scenarios. First, it is possible that the reported "similarity boost" on vocabulary size is not directly related to the amount of cognates shared between both languages, but rather to the overlap across their correspondent phonological inventories. One of the measures of language similarity Floccia et al. computed is the mean phonological Levenshtein distance across translation equivalents. This involves the calculation of the number of edit operations (e.g., insertions, deletions or replacements) the phonological transcription of a word-form has to go through to become identical to that of its translation equivalent. These scores were aggregated across pais of tanslation equivalents, obtaining a single similarity estimate of for each language pair. This measure embeds confounded information about the "degree of cognateness" and the amount of phonological overlap between both languages. Languages that are phonologically closer tend to share more cognates than those that are phonologically more distant (e.g., Spanish-Italian, Enlgish-Dutch), with few exceptions (e.g., Spanish-Basque). In the light of these data, it is not possible to discern whether vocabulary size is increased by the ealier acquisition of cognates or by the overlap across the phonological systems of the languages of a bilingual. We address this issue by testing two hypotesis.

## Hypothesis 1: Cognates are acquired earlier than non-cognates

First, we test whether the increase in vocabulary size reported by Floccia et al. for languages sharing a greater amount of cognates is indeed caused by cognates being acquired earlier. In line with this account, @bosch2014 found higher productive vocabulary scores in 18 month-old Catalan-Spanish bilinguals, compared to monolinguals, due to the higher presence of cognates their lexicon. We will adopt a similar approach to @braginsky2019, who modelled cross-sectional data from individual items to explore what properties were associated with an earlier acquisition. We will analyse the acquisition trajectories of a selection of `r dat_items %>% filter(include) %>% group_by(language) %>% summarise(n = n(), .groups = "drop") %>% pull(n) %>% max()` translation equivalents in a cross-section sample of Catalan/Spanish bilinguals aged 14 to 34 months of age. This approach allows to analyse how the proportion of infants that know each item changes across ages, providing a more suitable framework to put developmental hypotheses to test, instead of taking a "snapshot" of the state of the lexical of children at a single age.

## Hypothesis 2: Cognates are acquired closer in times than non-cognates

One mechanism that could acccount for an ealier acquisition of cognates is that parallel activation makes word-forms acquired in one language available during the acquisition of word forms in the other language. One consequence of this situation is that translation equivalents scaffold the acquisition of each other. This would point to different mechanisms engaged in the acquisition of cognates vs. non-cognates. While the phonology of cognate translation equivalents overlaps, and thus could be co-activated, the phonology of non-cognates does it in a much lesser degree. Consider the case of a Catalan-Spanish bilingual toddler that has already acquired the Catalan words *porta* (/ˈpɔr.tə/, door) and *taula* (/ˈtaw.ɫə/, table), but not yet their correspondent translation equivalents in Spanish (*puerta*, /ˈpwer.ta/; *mesa* /ˈme.sa/). According to our hypothesis, hearing the word *puerta* in presence of a door should lead to the activation of *porta*, which should facilitate the acquisition of the former. On the contrary, when hearing *mesa* in presence of a table, it is unlikely that the phonology of *taula* will be activated. This mechanism should lead to cognate translation equivalents being learnt closer in time than non-cognate translation equivalents. To test this hypothesis, we will extract the age of acquisition of each word-form in the questionnaire (as estimated by the model used to test Hypothesis 1), and will calculate the difference in time between the acquisition of such word-forms and their translation equivalents. We will test how cognateness affects such outcome.

@@

We collected data from `r dat_raw$id_db %>% unique %>% length() %>% printnum()` bilinguals from the Metropolitan Area of Barcelona, between `r min(dat_raw$time_stamp, na.rm = TRUE) %>% format(., "%dth %B, %Y")` and `r max(dat_raw$time_stamp, na.rm = TRUE) %>% format(., "%dth %B, %Y")`. All families participated voluntarily. This study was approved by the Comitè d'Ètica de la Investigació amb Medicaments (CEIm) from Hospital del Mar (Barcelona, Spain), code XXXXXXXXX.

We assessed toddlers' language background asking parents for an estimated proportion of exposure to each language. We defined participants as Spanish- or Catalan-dominant according to the language they were exposed the most to. Participants were considered bilingual - therefore included - if exposed 50-90% of the time to the non-dominant language. We excluded participants with >10% exposure to a third language. Fig. \@ref(fig:lpdist) shows how participants were distributed across age bins and language profiles.

```{r lpdist, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

dat_participants %>%
  filter(completed, lp %!in% "Other", age_bin %in% bins) %>%
  ggplot(aes(x = age_bin, fill = dominance)) +
  geom_bar() +
  labs(x = "Age (months)", y = "Number of partcipants",
       fill = "Dominant language") +
  scale_y_continuous(labels = as.integer) +
  scale_fill_brewer(palette = "Set1") +
  theme_custom +
  theme(legend.position = c(0.05, 0.85),
        legend.direction = "vertical",
        legend.justification = 0,
        panel.grid.major.y = element_line(colour = "grey", linetype = "dotted"))
```

## Questionnaire

Parents completed an online questionnaire that gathered information about their demographic status, language environment of the child, and that included two vocabulary inventories: one for Catalan and one for Spanish. We implemented the questionnaire in `formr` [@arslan2020], a free, open-source platform. Participants were randomly allocated into one of four list of items. Each list contained a distinct inventory of translation equivalents. We included items from a diverse sample of `r distinct(dat_items, category) %>% nrow()` semantic/funcional categories:  Action words, adventures, animals, body parts, clothes, color, descriptive words, food and drink, furniture and rooms, household items, on-line, outside, parts of animals, parts of things, people, pronouns, quantifiers, question words, time, toys, and vehicles. We discarded the following categories: adverbs, auxiliary words, connectives, interjections and games and routines. Those categories that did not entail more than 16 items - thus resulting in less than four items after dividing it in four lists - were not divided: all of their items were included in the four lists. Another subset of items that were part of the trial lsits of some experiments in the lab were also included in all versions.

Each item of the inventory presented a word-form, to which parents were instructed to report whether the toddlers was able to understand it, understand ans say it, or neither. The Catalan inventory contained `r filter(dat_items, language=="Catalan") %>% pull(item) %>% unique() %>% length()` items (`r filter(dat_items, cognate_rater1, language=="Catalan") %>% pull(item) %>% unique() %>% length()` cognates, `r filter(dat_items, !cognate_rater1, language=="Catalan") %>% pull(item) %>% unique() %>% length()` non-cognates) and the Spanish inventory contained `r filter(dat_items, language=="Spanish") %>% pull(item) %>% unique() %>% length()` (`r filter(dat_items, cognate_rater1, language=="Spanish") %>% pull(item) %>% unique() %>% length()` cognates, `r filter(dat_items, !cognate_rater1, language=="Spanish") %>% pull(item) %>% unique() %>% length()` non-cognates) . Items of one language were translation equivalents from the other (e.g., whenever *gos* [dog] was included in the Catalan inventory, the word *perro* was included in the Spanish inventory), roughly following a one-to-one mapping. When there were two acceptable translation equivalents for a given word, we included both (e.g., Catalan *acabar* [*to finish*] and Spanish *acabar* and *terminar*), or merged them into the same one (e.g., Spanish *mono* [*monkey*] and Catalan *mono/mico*). We excluded from the analysis multi-word items (e.g., *barrita de cereales* [cereal bar]) and items that included more than one word-form (e.g., *mono / mico*).

For testing Hypothesis 1, we modelled the proportion of infants that understood or produced each item  within age bins of 2 months. Our model assumes that this proportion follows a logistic curve accross ages, an that the age of acquisition of each item is generated from a linear model that adjusts for the dominance of the language of the item (*item dominance*, L1 vs. L2), the cognate status of the word (*cognateness*, Non-cognate vs. Cognate), and the frequency of the word form in its correspondent language. Our prediction  is that the age of acquisition will be lower for cognate words, compared to that of non-cognate words.

Items were classified into cognates and non-cognates by a trained Catalan-Spanish bilingual with broad experience in phonology and phonetics. Items were classified as L1 if the item was included in the inventory of the language the participant was dominant in, and as L2 if otherwise For instance, the item porta (door, in Catalan) was classified as L1 when answered by Catalan-dominant participants, and as L2 when answered by Spanish-dominant participants. We retrieved frequencies from SUBTLEX-ESP [@cuetos2011] and SUBTLEX-CAT [@boada2019] for Spanish and Catalan words, respectively. Word frequency was calculated as the Zipf score of the word [@zipf1949; @heuven2014]. Missing frequency scores were imputed using multiple imputation [@vanbuuren2011]. Table \@ref(tab:itemsinfo) shows the classification of items in cognates and non-cognates and their frequency scores across the four lists of the inventories.

```{r itemsinfo, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

dat_items %>%
  drop_na(cognate_rater1) %>%
  filter(include) %>%
  select(A, B, C, D, cognate_rater1, frequency, language) %>% 
  mutate(cognate_rater1 = ifelse(cognate_rater1, "Cognate", "Non-cognate")) %>%
  pivot_longer(c(A, B, C, D), names_to = c("list"), values_to = "n") %>% 
  filter(n) %>% 
  group_by(list, language, cognate_rater1) %>% 
  summarise(n = sum(n, na.rm = TRUE),
            mean = mean(frequency, na.rm = TRUE),
            sd = sd(frequency, na.rm = TRUE),
            min = min(frequency, na.rm = TRUE),
            max = max(frequency, na.rm = TRUE)) %>%
  ungroup() %>% 
  rename(cognateness = cognate_rater1) %>%
  relocate(list, language, cognateness, n, mean, sd, min, max) %>%
  arrange(list, language, cognateness) %>%
  mutate(list = paste("List", list)) %>% 
  pivot_wider(names_from = cognateness, values_from = c(n, mean, sd, min, max)) %>%
  clean_names() %>%
  gt(groupname_col = c("language")) %>% 
  fmt_number(columns = contains("cognate"), decimals = 2) %>%
  fmt_number(columns = starts_with("n_"), decimals = 0) %>%
  tab_header(title = md("**Questionnaire and item properties**"),
             subtitle = "Participants were randomly allocated on of the the four lists, and completed both the Catalan and the Spanish versions.") %>% 
  tab_spanner(label = "Frequency (Cognates)", columns = vars("n_cognate", "mean_cognate", "sd_cognate", "min_cognate", "max_cognate")) %>% 
    tab_spanner(label = "Frequency (Non-cognates)", columns = contains("non_cognate")) %>% 
  cols_label(list = md("**List**"), 
             n_cognate = md("*N*"),
             mean_cognate = md("*Mean*"),
             sd_cognate = md("*SD*"),
             min_cognate = md("*Min*"),
             max_cognate = md("*Max*"),
             n_non_cognate = md("*N*"),
             mean_non_cognate = md("*Mean*"),
             sd_non_cognate = md("*SD*"),
             min_non_cognate = md("*Min*"),
             max_non_cognate = md("*Max*")) %>%
  gtsave(here("Results", "descriptives_items.png"))

@@

We fit two separate logistic models on comprehensive and productive data, respectively. Logistic curves are characterised by three parameters[see @mahr2019 and @mahr2020 for an excellent introduction to this approach, in which we based this analysis]: (a) an *asymptote* (`asym`, upper boundary  of the proportion of infants that have acquired the item), (b) the *steepness* (`steep`, how fast the a given item is acquired by infants), and (c) a *mid-point* (`mid`, the age at which steepness is maximum). We work on the assumption that the trajectory of aquisition of items follows a sigmoidal shape [@mayor2011], and that the age of acquisition of a word is fairly captured by the position of the mid-point of its curve. This approach allows us to model and estimate the age of acquisition of items explicitely. Our model estimated the asymptote and stepness as grand means across items (i.e., intercepts). Mid-points were estimated by a linear zero-one-beta-regression model^[Zero-or-one-inflated beta regression models allow to account for overdispersion then using proportions as the output of a model accounting, while also allowing values to be at boundary (0 and 1).][@ospina2012] that included our predictors of interest: `age`, `dominance` (contrast-coded, L2 = -0.5, L2 = 0.5), `cognateness` (contrast-coded, Non-cognate = -0.5, Cognate = 0.5), their interaction, and `frequency` (standardised Zipf score). We also included a random intercept for each translation equivalent (`TE`). Appendix 1 shows a detailed description of the model.

We estimated the parameters of our model using the Bayesian framework. Bayesian models use the Bayes rule to combine prior knowlegde about the distribution of a given parameter, and the likelihood of each of the values it can take given tha observed data. We relied on the available data on Wordbank [@frank2017] to generate our priors (see Appendix 1 for more details about our priors). In order to test the contribution of the predictor *cognateness* on the fit of the model,  we compared the fit of our (expanded) model to the fit of a (null) model that did not include *cognateness* as predictor [@schad2020] using leave-one-out cross validation. We fit both the null and alternative models using the R environment [@rcoreteam2013; @rstudioteam2015] and the `brms` package [@burkner2017], which relies on the probabilistic language Stan [@carpenter2017] to approximate posterior distributions. We used Pareto-smoothed importance sampling [PSIS; @vehtari2017; @vehtari2019] to compare the null and extended models. Data and model results were processed and visualised using the `tidyverse` family of R packages [@wickham2019] and the `tidybayes` package [@kay2020].

To test Hypothesis 2, we run a Bayesian ANOVA using the difference in mid-points between the two word-forms of each translation equivalent (in Catalan and Spanish, respectively), that included *cognateness* as predictor. Using the `BayesFactor` R package [@morey2018a], we computed a Bayes factor (*BF*) that compared the likelihood of a linear model that included *cognateness* as a predictor against the likelihood of a linear model that did not, under the light of our data.

## Hypothesis 1

### Comprehensive vocabulary

### Productive vocabulary

## Hypothesis 2

### Comprehensive vocabulary

### Productive vocabulary

## Appendix 1

Model formula and priors:

$$proportion_{ij} \sim \frac{asym}{1 + e^{(mid_i-age) \times steep}} \\ asym \sim Normal(0.79, 0.1) \\ steep \sim Normal(1.76, 0.8) \\mid \sim Beta(\mu_i, \phi) \\ \mu_i = \beta_0 + TE_{0i} + \beta_1 \times dominance_i + \beta_2 \times cognateness_i \\ \beta_0 \sim Normal(4.50, 1) \\ \beta_1 \sim Normal(0, 10) \\ \beta_2 \sim Normal(0, 10) \\\phi \sim Normal(1.5, 1)$$

## Appendix 2: Computing priors
